---
title: "Practice Midterm Exam 2"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2: Data Analysis

The data analyzed in this part of the exam came from medical patients who have suffered from a heart attack within the past year. 

The response variable is *still_alive* which is a binary variable indicating if the patient is still living or not (0=dead, 1=alive).

The predicting variables are as follows:

* age -- age in years when heart attack occurred
* pericardial_effusion-- Pericardial effusion is fluid around the heart. This is a binary variable: 0=no fluid, 1=fluid.
* fractional_shortening -- a measure of contractility around the heart. Lower numbers are increasingly abnormal
* lvdd -- left ventricular end-diastolic dimension. This is a measure of the size of the heart at end-diastole. Large hearts tend to be sick hearts

Read the data and answer the questions below:
```{r}
# Loading of the data
echo= read.csv("echo.csv", header=TRUE, sep=",")

# Viewing first few rows of data
head(echo)
```
## Q21
Plot the boxplots to describe the relationship between the response (still_alive) and the following predicting variables: *age*,  *fractional_shortening*, and *lvdd*. Interpret the plots.

**Response to Q21**:
***the median values for alive and dead are not equal for any of the three predicting variables. For each of the predicting variables, there appears to be a relationship with alive and dead. Reasonable, we could use this to predict the probability of alive/dead. Logistic regression is reasonable.

```{r}

par(mfrow=c(2,2))
boxplot(echo$age ~ echo$still_alive, xlab = "age", ylab = "alive")
boxplot(echo$fractional_shortening ~ echo$still_alive, xlab = "fractional_shortening", ylab = "alive")
boxplot(echo$lvdd ~ echo$still_alive, xlab = "lydd", ylab = "alive")
```

## Q22
Create a logistic regression model using the following predicting variables: *age*, *fractional_shortening*, and *lvdd*.  

a) What are the model parameters and what are their estimates? 

**Response to Q22a):**
(Intercept)  8.54944187                   
echo$age: -0.08983858 
echo$fractional_shortening: 5.90295498                  
echo$lvdd:-0.65886388 

b) Which of the parameters are statistically significant at the 95% confidence level? 

**Response to Q22b):**
***All of the parameters are significant at the 95% confidence level.

```{r}
mod1 <- glm(echo$still_alive ~ echo$age + echo$fractional_shortening + echo$lvdd, family = binomial(link = "logit"))

summary(mod1)

mod1$coefficients

p_values = summary(mod1)$coef[,4]
alpha = 0.05
p_values <= alpha
```
## Q23
Provide the equation for the estimated logit transformation of the probability of still being alive given the predicting variables. 

**Response to Q23:**
$$ logit = log(\frac{p(x)}{1-p(x)}) = 8.54 + -.089{Xage,i} + 5.90{Xfrac,i} - .658{Xlvdd,i} $$
***MISC EQAUTION FOR PROBABILITY***
$$p =  \frac{e^{B0+B1(-0.08) + B2(5.90) + B3(-0.65)}}{1 + e^{B0 + B1(-0.08) + B2(5.90) + B3(-0.65)}} $$
## Q24
Interpret the estimated value of the parameter corresponding to *age* in the context of this problem. 

**Response to Q24:** for a one unit increase in age, the log-odds of still_alive decreases by -0.089, holding all other variables constant. Similarly, the odds changes by a factor of $$e^{-.089} = .9414$$, holding all other variables constant. 


## Q25
Perform a test for the overall regression, using $\alpha = 0.05$. Does the overall regression have explanatory power?

**Response to Q25:**
H0: B1 = B2 = B3 = 0
Ha: at least one of the coefficient !=0

Given that the p-value, 1.474676e-05, < alpha of 0.05, we reject the null hypothesis. At least one of the coefficients is nonzero. The model has explanatory power. 

```{r}
testStat <- mod1$null.deviance - deviance(mod1)
pValue <- 1-pchisq(testStat,length(coef(mod1)) -1)
cbind(testStat, pValue)
summary(mod1)
```

## Q26
How does the log-odds change with a 1 unit increase in *fractional_shortening*, holding all other variables constant? Explain.

**Response to Q26** with a one unit increase in fractional_shortening, the log-odds of still alive increases by 5.90, when all else constant. Similarly, the odds changes by a factor of $$e^{5.90} = 365$$, when all else constant.

## Q27
What is the approximated distribution of the null deviance? What is the value of the null deviance and its degree of freedom?

**Response to Q27:** the ~ distribution of the null deviance is chi-sqaured. 

```{r}
#deviance
mod1$null.deviance

#deviance df
dim(echo)[1]-1
```





**The End.**
