---
title: "Practice Midterm Exam 1"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 2: Data Analysis

For this part, you will need to use R and the provided dataset.

The dataset was collected from Airbnb with data on listings in the city of Asheville, NC.

Here is the data provided for each listing:

* room id: A unique number identifying an Airbnb listing.
* host id: A unique number identifying an Airbnb host.
* room type: One of 'Entire home/apt','Private room', or 'Shared room'
* reviews: The number of reviews that a listing has received.
* overall satisfaction: The average rating (out of five) that the listing has received from those visitors who left a review.
* accommodates: The number of gusts a listing can accommodate.
* bedrooms: The number of bedrooms a listing offers.
* price: The price (in USD) for a night stay. In early surveys, there may be some values that were recorded by month.


To read the data in R, save the file in your working directory (make sure you have changed the directory if different from the R working directory) and read the data using the R function read.csv().

```{r}
# Read in the data
house = read.csv("tomslee_airbnb_asheville_1498_2017-07-20.csv", head = TRUE, sep = ",")
# Show the first few rows of data
head(house, 3)

```

## Question 25

Fit a multiple linear regression model named 'model1' using price as the response variable and the following predicting variables: room type, reviews, overall satisfaction, accommodates, and bedrooms. 

Which coefficients (including intercept) are statistically significant at the 99% confidence level?

```{r}
model1 <- lm(price ~ room_type + reviews + overall_satisfaction + accommodates + bedrooms, data = house)

summary(model1)
confint(model1,level = 0.99)
```
**Response to Question 25**: based on the 99% confidence interval the following do not span 0 and therefore are significant at the 99% level: intercept, room type private, room type shared, overall satisfaction, accommodates, and bedroom.

## Question 26

What is the estimated coefficient for room type = "Private Room"?

```{r}

model1$coefficients["room_typePrivate room"]
```
**Response to Question 26**: -33.28108

## Question 27

What is the interpretation for the estimated coefficient for room type = "Private Room"?

```{r}
summary(model1)

```
**Response to Question 27**: A private room has a listing price -$33.00 in contrast to base case (room_type Entire Home/Apt), holding all else constant.


## Question 28

Assess whether the model assumptions hold.

```{r}
plot(model1)
hist(model1$residuals,main="Histogram of Residuals")

```
**Response to Question 28**: 

*Linearity/Mean Zero: Okay. appears to hold reasonably well although as you move along the x axis data points do appear to be increasing slightly

*Independence: No. errors appear to have some correlation and follow a pattern as they are closely clustering near lower fitted values. 

*Constant Variance: No. Errors appear to be increasing along the x-axis (spread/hederoscadacity)

*Normality: No. Based on the Normal QQ plot and histogram, normality assumption not satisfied. Data skewed to right.


## Question 29

Using your model, make a prediction for a listing on Airbnb in Asheville with the following factors: bedrooms = 1, accommodates = 2, reviews = 92, overall_satisfaction = 3.5, and room_type = 'Private room'. What is your predicted price for such a listing and the corresponding 95% prediction interval?

```{r}

newdata1 <- data.frame(bedrooms = 1, accommodates = 2, reviews = 92, overall_satisfaction = 3.5, room_type = "Private room")
  
predict(model1, newdata = newdata1, interval = "prediction", level = 0.95)

```
**Response to Question 29**: predicted price = $72.58



**The End.**